Alternative Methods for Retaining Explicit and Finding Implicit Sharing in Embedded DSLs

### 2023-02-14 paper/paper.tex
updated figures

  Kiselyov's methods for implicit and explicit sharing detection for finally
  tagless style DSLs is an elegant solution without having to resort unsafe
  observable sharing. However these methods are not applicable to all types of
  DSLs (including those generating hypergraphs). We will present alternative
  methods which handle these cases. The main difference comes from the use of a
  trie to perform hashconsing. Our method for implicit sharing essentially
  trades worst-case exponential growth in computation for increased memory
  footprint. To mitigate this issue, our method for explicit sharing reduces the
  memory footprint.

@@

instance Exp Pretty where
  add x y = Pretty $ "("++unPretty x++") + ("++unPretty y++")"
  variable x = Pretty x
  constant x = Pretty $ show x
\end{minted}

@@

In order to generate such a data structure, we will need to keep track of the
current maximum identifier to keep them unique. The representation for the
finally tagless instance is then a wrapper around a state monad that holds the
DAG being constructed in its state and returns the current (top)
\mintinline{haskell}{NodeID}:

@@

The technique is essentially that of hash-consing, popularized by its use in
LISP compilers, but discovered by Ershov in 1958 \cite{ershov1958:consing}.
Other works have explored the use of type-safe hash-consing in embedded DSLs,
see \cite{filliatre:typesafeconsing}.

@@

The heart of our problem is that whenever we need to sequence the state of the inputs
for one of our DSL functions we want to first check if it's already been
evaluated. But how do we do that without first evaluating it to gain access to
its unique identifier? We need some way to uniquely identify it outside the monad.

@@

To prevent confusion, we name the hash-consing function in our method \mintinline{haskell}{triecons}:
\begin{minted}{haskell}
triecons :: ByteString -> Node -> State DAG NodeID
triecons sAST node = do
  DAG trie maxID <- get
  case Trie.lookup sAST trie of
    Nothing -> let maxID' = maxID+1
                   trie' = Trie.insert sAST (node,maxID') trie
                in do put $ DAG trie' maxID'
                      return maxID'
    Just (_,nodeID) -> return nodeID
\end{minted}
We use it to implement the DAG-building instance of the DSL,
which looks a lot like the previous instance.
The substantial differences are the 
\mintinline{haskell}{buildStringAST} calls
which you can think of as pretty printing, but optimized for the trie,
and the use of \mintinline{haskell}{seqArgs} (explained below):
\begin{minted}{haskell}
instance Exp Graph where
  constant x = let
    node = NConstant x
    sAST = buildStringAST node []
    in Graph (triecons sAST $ NConstant x) sAST
  variable x = let
    node = NVariable x
    sAST = buildStringAST node []
    in Graph (triecons sAST $ NVariable x) sAST
  add e1 e2 = let
      sAST = buildStringAST "nadd" [e1,e2]
      sT = do ns <- seqArgs [e1,e2]
              case ns of
                [n1,n2] -> triecons sAST $ NAdd n1 n2
                _ -> error "black magic"
    in Graph sT sAST
\end{minted}
The magic is in \mintinline{haskell}{seqArgs}.
We only evaluate the inner state \mintinline{haskell}{sT} of each argument if we
fail to find its corresponding serialized AST in the Trie.
\begin{minted}{haskell}
seqArgs :: [Graph a] -> State DAG [NodeID]
seqArgs inps =
  let
    seqArg (Graph sT sAST) =
      do DAG trie _ <- get
         case Trie.lookup sAST  trie of
           Nothing -> sT
           Just (_,nodeID) -> return nodeID
  in sequence $ map seqArg inps
\end{minted}

@@

\begin{figure}[!h]
  \centering
  \includegraphics[width=0.45\textwidth]{figs/noexplicit.png}
  \hfill
  \includegraphics[width=0.45\textwidth]{figs/explicit.png}
  \caption{Comparison of cos using hashcons vs triecons}
  \label{fig:explicitvsno}
\end{figure}

@@

The method has its drawbacks in terms of memory usage, but these can be
mitigated by explicitly specifying sharing. This does present an extra burden on
the DSL writer to implement explicit sharing when necessary and ensure labels
are not reused. Future work may investigate the use of a preprocessor or plugin
to automate explicit sharing.

>>>>

  Kiselyov's methods for implicit and explicit sharing detection for finally
  tagless style DSLs is an elegant solution without having to resort to unsafe
  observable sharing. However these methods are not applicable to all types of
  DSLs (including those generating hypergraphs). We will present alternative
  methods which handle these cases. The main difference comes from the use of a
  trie to perform hashconsing. Our method for implicit sharing essentially
  trades worst-case exponential growth in computation for increased memory
  footprint. To mitigate this issue, our method for explicit sharing reduces the
  memory footprint.

@@

instance Exp Pretty where
  add x y = Pretty $ "("++runPretty x++") + ("++runPretty y++")"
  variable x = Pretty x
  constant x = Pretty $ show x
\end{minted}

@@

In order to generate such a data structure, we will need to keep track of the
current maximum identifier to keep them unique. The representation for the
finally tagless instance is then a wrapper around a state monad that holds the
DAG being constructed in it's state and returns the current (top)
\mintinline{haskell}{NodeID}:

@@

The technique is essentially that of hash-consing, popularized by it's use in
LISP compilers, but discovered by Ershov in 1958 \cite{ershov1958:consing}.
Other works have explored the use of type-safe hash-consing in embedded DSLs,
see \cite{filliatre:typesafeconsing}.

@@

The heart of our problem is that whenever we need to sequence the state of the inputs
for one of our DSL functions we want to first check if it's already been
evaluated. But how do we do that without first evaluating it to gain access to
it's unique identifier? We need some way to uniquely identify it outside the monad.

@@

To prevent confusion, we name the hash-consing function in our method \mintinline{haskell}{triecons}:
\begin{minted}{haskell}
triecons :: ByteString -> Node -> State DAG NodeID
triecons sAST node = do
  DAG trie maxID <- get
  case Trie.lookup sAST trie of
    Nothing -> let maxID' = maxID+1
                   trie' = Trie.insert sAST (node,maxID') trie
                in do put $ DAG trie' maxID'
                      return maxID'
    Just (_,nodeID) -> return nodeID
\end{minted}
We use it to implement the DAG-building instance of the DSL,
which looks a lot like the previous instance.
The substantial differences are the 
\mintinline{haskell}{buildStringAST} calls
which you can think of as pretty printing, but optimized for the trie,
and the use of \mintinline{haskell}{seqArgs} (explained below):
\begin{minted}{haskell}
instance Exp Graph where
  constant x = let
    node = NConstant x
    sAST = buildStringAST node []
    in Graph (triecons sAST $ NConstant x) sAST
  variable x = let
    node = NVariable x
    sAST = buildStringAST node []
    in Graph (triecons sAST $ NVariable x) sAST
  add e1 e2 = let
      sAST = buildStringAST "nadd" [e1,e2]
      sT = do ns <- seqArgs [e1,e2]
              case ns of
                [n1,n2] -> triecons sAST $ NAdd n1 n2
                _ -> error "black magic"
    in Graph sT sAST
\end{minted}
The magic is in \mintinline{haskell}{seqArgs}.
We only evaluate the inner state \mintinline{haskell}{sT} of each argument if we
fail to find it's corresponding serialized AST in the Trie.
\begin{minted}{haskell}
seqArgs :: [Graph a] -> State DAG [NodeID]
seqArgs inps =
  let
    seqArg (Graph sT sAST) =
      do DAG trie _ <- get
         case Trie.lookup sAST  trie of
           Nothing -> sT
           Just (_,nodeID) -> return nodeID
  in sequence $ map seqArg inps
\end{minted}

@@

\begin{figure}[!h]
  \centering
  \includegraphics[width=0.45\textwidth]{figs/noexplicit.png}
  \hfill
  \includegraphics[width=0.45\textwidth]{figs/explicit.png}
  \caption{Comparison of tan using no explicit sharing vs explicit}
  \label{fig:explicitvsno}
\end{figure}

@@

The method has it's drawbacks in terms of memory usage, but these can be
mitigated by explicitly specifying sharing. This does present an extra burden on
the DSL writer to implement explicit sharing when necessary and ensure labels
are not reused. Future work may investigate the use of a preprocessor or plugin
to automate explicit sharing.
